---
title: "Query & model preview"
subtitle: "A Shiny workbench where an LLM writes SQL and model specs, R executes them, and every intermediate is on screen. A technical walkthrough."
date: last-modified
date-format: "MMMM D, YYYY"
format:
  html:
    theme: [cosmo, styles.scss]
    toc: true
    toc-depth: 3
    toc-title: "On this page"
    toc-location: left
    page-layout: article
    anchor-sections: true
    smooth-scroll: true
    code-copy: true
    link-external-newwindow: true
    fig-cap-location: bottom
execute:
  echo: false
  warning: false
---

```{r setup}
#| include: false

# Screenshot slots. A missing file renders a dashed placeholder instead of a
# broken image, so the page still builds mid-edit.
shot <- function(file, caption, alt = caption) {
  path <- file.path("images", file)
  if (file.exists(path)) {
    knitr::asis_output(sprintf(
      '<figure class="shot"><img src="%s" alt="%s"><figcaption>%s</figcaption></figure>',
      path, htmltools::htmlEscape(alt, attribute = TRUE), caption
    ))
  } else {
    knitr::asis_output(sprintf(
      paste0('<div class="shot-slot"><span class="slot-label">Screenshot</span>',
             '<span class="slot-file">docs/images/%s</span>',
             '<span class="slot-note">%s</span></div>'),
      file, caption
    ))
  }
}
```

::: {.hero}
::: {.hero-lede}
The model does not analyze your data. It emits a **typed tool call** — SQL plus a
plot or model spec — and R does everything after that.
:::

That split is the whole design. The LLM's output is a structured argument object
validated against a schema, not prose and not code that gets `eval()`ed. R owns
execution, the database, the rendering, and the statistics. What reaches the model
afterwards is the result R computed.

Every screenshot below is a live session against `midwest` — 437 Midwestern
counties, 28 columns, from `ggplot2`. Nothing in the app was tuned for it.

::: {.hero-meta}
[R + Shiny]{} [ellmer]{} [querychat]{} [DuckDB]{} [tidymodels]{} [ggplot2]{} [Claude Haiku 4.5]{}
:::

Live at [ellmer-practice.ethandbard.com](https://ellmer-practice.ethandbard.com), behind Cloudflare Access on Ethan's own API key. Everyone else runs it locally with their own key — see the [README](https://github.com/ethandbard/ellmer-practice#readme).
:::

## Architecture

One turn is one tool call. There is no agent loop, no multi-step planning, and no
path by which the model reaches the database directly.

```{mermaid}
flowchart TB
  subgraph LLM["Model — Anthropic or xAI"]
    T["Emits one typed tool call<br/><b>set_dashboard</b> or <b>set_model</b>"]
  end
  subgraph RS["R — Shiny process"]
    G["<b>is_read_only_sql()</b><br/>reject non-SELECT"]
    D[("DuckDB<br/>table <b>data</b>")]
    RC["Plot recipe /<br/>tidymodels workflow"]
    P["Panes: chart · SQL · code"]
    S["Result summary<br/>+ profile"]
  end
  Q["Your question"] --> T
  T --> G
  G -->|"SELECT / WITH only"| D
  D --> RC
  RC --> P
  RC --> S
  S -->|"numbers R computed"| T
  G -.->|"rejected: error text<br/>names allowed columns"| T
```

The dotted edge matters as much as the solid ones: a failed query returns a message
naming the legal columns and asking for a retry, so a hallucinated column name
becomes a self-correcting round trip rather than a broken pane.

## What the model is actually given

### A statistical profile, not rows

By default the chat client never receives the table contents. On load, R profiles
every column and writes that profile into the system prompt. A `describe_data`
tool lets the model re-read it on demand.

```{r}
shot("profile-greeting.jpg",
     "The session opening on <code>midwest</code>: 437 rows, 28 columns, and the full column list — the model's entire view of the table before it asks anything.")
```

Roles are inferred from **statistics**, with column names used only as a small
tiebreak. From `R/profile.R`:

```r
profile_roles <- c("measure", "count", "category", "clock", "id", "text", "constant")
```

The classifier reads cardinality, type, integer-ness, sign, and missingness. A few
of the rules that make it survive unfamiliar tables:

```r
if (d <= 1) return("constant")                      # single distinct value
if (isTRUE(row$parsed_clock)) return("clock")       # parses as a date

# character/factor with a small, stable set of levels
if ((is.character(x) || is.factor(x)) && d >= 2 && d <= 20) return("category")

# count-named non-negative integers beat the small-integer category rule:
# food_delivery$items is 1-8 and is a count, not a grouping
if (is.numeric(x) && isTRUE(row$is_integerish) && isTRUE(row$all_nonneg) && count_name) {
  # ... unless it is a 1-10 rating/score, which is a measure
  return("count")
}

# numeric encodings of categories: cyl, gear, carb
if (isTRUE(row$is_integerish) && !yearish && d >= 2 && d <= 12 && ratio < 0.15) {
  return("category")
}
```

Integer years are explicitly *not* treated as a clock, and a 1–10 column named
`rating` is a measure rather than a count. These are the cases that break
name-matching schemas.

The same profile feeds four consumers — the system prompt, `auto_view()` (the
default chart), `auto_features()` (default model predictors), and the
`describe_data` tool — so the app's automatic choices and the model's suggestions
are derived from one description of the data.

### A typed tool schema

`set_dashboard` is an `ellmer` schema, not a prompt convention. Every field is
optional and enum-constrained where possible, and the enums are generated from the
same registries the renderer uses — adding a theme is one list entry, not a second
schema edit.

```r
plot_spec_type <- function() {
  ellmer::type_object(
    .description = "What to show. Data mapping for the featured ggplot.",
    from   = opt_enum(c("detail", "aggregate", "auto"), "Which SQL result to plot."),
    geom   = opt_enum(c(allowed_geoms, "auto"), "ggplot geom. bar is accepted as col."),
    x      = opt_string("Column for x, from the chosen SQL result."),
    y      = opt_string("Column for y. Omit for histogram/density."),
    color  = opt_string("Optional color/fill column."),
    facet  = opt_string("Facet wrap column, or the grid's row variable."),
    smooth = opt_enum(c("none", "lm", "loess"), "Add a smooth to point/jitter charts."),
    # ... sort, position, titles, annotations, reference lines, bounds
    .required = FALSE
  )
}
```

The geom enum is closed:

```r
allowed_geoms <- c(
  "col", "line", "point", "histogram", "boxplot", "density", "area",
  "violin", "jitter", "tile", "hex", "bin2d", "ecdf", "qq",
  "errorbar", "pointrange", "step", "ribbon"
)
```

**Style is a separate, sticky channel.** `style_spec_type()` carries theme,
palette, accent, highlight, legend position, scale transforms, number formats and
label sizes. Merge semantics are partial-update: only the fields sent are
overwritten, and they persist until reset.

That is what makes *"now make it dark"* a one-field call. The model omits both SQL
strings and the entire plot mapping; R reuses the last result set and re-renders.
From `app.R`:

```r
reused <- reuse_last_queries(detail, agg, list(
  detail_df     = isolate(dash$detail_df),
  aggregate_df  = isolate(dash$aggregate_df),
  detail_sql    = isolate(dash$detail_sql),
  aggregate_sql = isolate(dash$aggregate_sql)
))
if (!detail$used && !agg$used) {
  return("No chart to restyle yet. Ask a data question first, or provide SQL.")
}
```

### SQL it is allowed to run

Model-authored SQL is gated before it reaches DuckDB. The interesting part is the
order of operations — comments and string literals are stripped *first*, so
`WHERE note = 'create'` is a legal read rather than a false positive:

```r
strip_sql_noise <- function(sql) {
  stripped <- gsub("--[^\\n]*", " ", sql)
  stripped <- gsub("/\\*.*?\\*/", " ", stripped)
  stripped <- gsub("'([^']|'')*'", "''", stripped)   # single-quoted literals
  stripped <- gsub('"([^"]|"")*"', '""', stripped)   # double-quoted identifiers
  trimws(stripped)
}

is_read_only_sql <- function(sql) {
  stripped <- strip_sql_noise(sql)
  if (!grepl("^(with|select)\\b", stripped, ignore.case = TRUE)) return(FALSE)

  # a second statement after a SELECT: `SELECT 1; DROP TABLE data`
  rest <- sub("^(with|select)\\b", "", stripped, ignore.case = TRUE)
  if (grepl(";", rest)) return(FALSE)

  !grepl("\\b(insert|update|delete|drop|alter|create|attach|copy|pragma|grant|truncate)\\b",
         stripped, ignore.case = TRUE)
}
```

::: {.callout-note appearance="simple"}
A source comment records why this is a keyword gate rather than a read-only
connection: on Windows, DuckDB cannot `ATTACH` the same file while a writer is
open, and an in-process `read_only = TRUE` handle still accepted `DROP` in testing.
:::

## Explore: one question, three synced panes

The Explore page opens with a chart nobody asked for — `auto_view()` picks it from
the profile, so the page is never empty and the model has something to describe.

```{r}
shot("explore-load.jpg",
     "<code>midwest</code> on load. <code>auto_view()</code> chose popdensity against area colored by state; the SQL and ggplot2 panes are already populated. Chat left, chart centre, code right — the split is drag-resizable.")
```

Then a question. *"Update the dashboard: average percent college-educated by state,
as a bar chart"* fires exactly one `set_dashboard` call:

```{r}
shot("explore-bar.jpg",
     "One tool call rewrites every pane. The chat answer quotes WI 20.02%, MI 19.42%, IL 18.79% — the same numbers you can read off the bars, because both came from the same result set.")
```

Nothing mapped "college-educated" to a column by hand. `percollege` was found in
the profile.

```{r}
shot("tool-call.jpg",
     "The transcript showing the <code>SET_DASHBOARD()</code> call itself. Expanding it reveals the exact arguments the model emitted.")
```

### The single-recipe invariant

This is the claim worth being precise about. The chart pane and the code pane are
not generated independently — they are two renderings of one plot recipe. The code
on screen is the code that ran.

::: {.panes}
```{r}
shot("pane-chart.jpg", "The rendered chart.")
```

```{r}
shot("pane-sql.jpg", "Pane 1 — the SQL the model wrote, formatted by <code>format_sql()</code>.")
```

```{r}
shot("pane-ggplot.jpg", "Pane 2 — the ggplot2 call generated from the same recipe that drew the chart above.")
```
:::

The SQL is the model's, reformatted for reading:

```sql
SELECT
  state,
  AVG(percollege) AS avg_college_percent
FROM data
GROUP BY state
ORDER BY avg_college_percent DESC;
```

The ggplot2 pane is generated, not transcribed — paste it into a script and you get
the same figure back:

```r
ggplot(aggregate, aes(x = state, y = avg_college_percent, fill = state)) +
  geom_col(position = "stack") +
  labs(title = "Average College-Educated Adults by State",
       x = "State", y = "Average % College-Educated") +
  theme_minimal(base_size = 13) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73",
                               "#F0E442", "#0072B2")) +
  theme(legend.position = "right")
```

The palette is Okabe–Ito. That is the default, not a request.

### Asking for a relationship rather than a chart type

*"Scatter of percent college-educated vs percent below poverty, colored by state,
with a linear smooth"* — one sentence, three schema fields (`geom = "point"`,
`color = "state"`, `smooth = "lm"`).

```{r}
shot("scatter-smooth.jpg",
     "<code>geom_point()</code> + <code>geom_smooth(method = \"lm\")</code> fit per state. More college-educated adults, less poverty — a negative slope in all five states, and the per-state fits differ enough to be worth noticing.")
```

Because style is sticky, the Okabe–Ito palette and theme carried over from the
previous question without being restated. Flipping **Interactive (Plotly)** in the
sidebar renders the same recipe through `plotly` with hover and zoom.

## Model: preview fits with a named baseline

The Model page fits a small `tidymodels` workflow and reports holdout metrics
**against a named baseline**. The baseline is the point — a metric without one is
not interpretable, and this app refuses to show one alone.

`set_model` accepts these methods:

```r
allowed_ml_methods <- c(
  "regression", "classification", "forecast",
  "poisson", "tree", "forest", "ridge", "elastic", "gam",
  "test", "pca", "kmeans", "correlation"
)
```

and these diagnostic plots:

```r
allowed_ml_plots <- c(
  "coef", "actual_pred", "residual", "forecast", "cv", "subgroup",
  "roc", "pr", "confusion", "calibration",
  "tree", "importance", "pdp", "path",
  "scree", "biplot", "cluster", "elbow",
  "heatmap", "stl", "test"
)
```

Method names are aliased (`lm`, `linear`, `ols`, `regress` → `regression`;
`glm`, `logistic` → `classification`), so the model does not have to guess the
canonical token.

### Regression, and a result that admits it is weak

*"Predict percent below poverty from percent college-educated, population density,
and metro status."*

```{r}
shot("model-regression.jpg",
     "Fit plot and coefficient estimates on the left; the model spec and the generated R on the right. The spec pane reports the split, the holdout metrics, and the baseline together.")
```

The spec pane reads:

```
method    regression
target    percbelowpoverty
features  percollege, popdensity, inmetro
n         437
split     random 80/20 holdout (seed 1) · train 350 / test 87

Holdout
rmse      4.567
mae       3.617
mape      32.9%
r2        -0.001
baseline  4.579 RMSE (train mean)
```

RMSE 4.567 against a train-mean baseline of 4.579, and R² ≈ 0. **The model barely
beats guessing the average, and the pane says so.** Reported alone, R² near zero
would look like a bug; next to a baseline it reads correctly as a weak predictor.

The coefficients are still real and still interpretable: each additional percentage
point of college attainment is associated with about **0.22 points less poverty**
(p < 0.001), and metro counties run **2.6 points lower** than rural ones. Modest
predictive power and genuine associations are not the same claim, and separating
them is exactly what the baseline buys you.

Cross-validation runs on training folds only. The R pane prints the actual call:

```r
frame <- data
# 80/20 split -> train, test
rec <- recipe(percbelowpoverty ~ percollege + popdensity + inmetro, data = train) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors())
spec <- linear_reg() |> set_engine("lm")
wf   <- workflow() |> add_recipe(rec) |> add_model(spec)
fit  <- fit(wf, data = train)
pred <- predict(fit, new_data = test)
tidy(extract_fit_parsnip(fit), conf.int = TRUE)
```

### Classification

*"Classify whether a county is in a metro area, and show the ROC curve."*

```{r}
shot("classification-metrics.jpg",
     "Logistic GLM via parsnip. Accuracy 0.875 against a 0.669 always-negative baseline, AUC 0.937, log-loss 0.321 — and the positive class stated explicitly as <code>1 vs 0</code>.")
```

Accuracy alone would be misleading here: the classes are unbalanced enough that
always guessing non-metro scores 0.669. The gap from 0.669 to 0.875, plus AUC
0.937 and cross-validated AUC 0.933 ± 0.005 on the training folds, is the real
result.

```{r}
shot("roc-curve.jpg",
     "The ROC pane for the fit above. See the note below — this curve is drawn inverted relative to the reported AUC.")
```

::: {.callout-warning appearance="simple"}
## Known issue: the ROC curve is drawn with the positive class flipped

The fit reports AUC **0.937**, but the plotted curve falls below the diagonal —
an area of roughly 0.063, which is 1 − 0.937. The metric is computed against
`positive 1 vs 0`; the curve is being traced against the other class. The numbers
in the metrics pane are correct, and the diagnostic plot is not. Left visible here
because it is a good illustration of why the app prints its metrics and its call
next to every figure — the disagreement is detectable precisely because both are
on screen.
:::

### A model you can read out loud

*"Show me a decision tree for metro status."*

```{r}
shot("decision-tree.jpg",
     "An <code>rpart</code> tree, depth-capped at 4 so the preview stays legible. Population density splits first (importance 75.4), then poverty rate.")
```

CV accuracy 0.868 against the logistic model's 0.875; CV AUC 0.864 against 0.937.
The tree is slightly worse and dramatically easier to explain — and because both
fits report the same metrics, the size of that trade is visible rather than
asserted.

### Unsupervised, same tool

*"Cluster the counties on their numeric columns."* — 25 numeric columns, k chosen
by the model at 3, projected onto the first two principal components.

```{r}
shot("kmeans-pca.jpg",
     "k-means on the first two PCs. One dense cluster of small rural counties, a broad middle, and a single far outlier — Cook County, IL.")
```

PCA, k-means and the correlation heatmap all route through `set_model`. When a
view has no coefficient table, the estimates pane says so rather than rendering an
empty chart.

### "Is the difference real?" is a first-class question

Significance questions dispatch to a test rather than a predictor. The choice among
t-test, ANOVA, chi-square and correlation comes from the column roles in the
profile.

```{r}
shot("correlation-test.jpg",
     "<code>Pearson cor.test</code> on the full sample: estimate −0.311, 95% CI [−0.394, −0.224], p < 0.001 — with the caveat printed by R, not written by the model.")
```

Metro counties run roughly 3–4 percentage points lower on poverty and zero is
nowhere near the interval. The result pane carries its own caveat — *correlation is
an association, not a cause* — emitted by R as part of the test output, so it
cannot be omitted by a model in a hurry.

## State, sessions, and the provider swap

Three implementation details that are easy to get wrong and are worth reading the
source for.

### The DuckDB table is process-wide

There is one table named `data` for the whole Shiny process. A session's
`reactiveValues` used to be seeded with the boot-time starter table regardless of
what a previous session had loaded — so a reload would claim `orders.csv` while the
real table held something else. The fix is a single process-level source of truth
that every session reads at start:

```r
current_source <- local({
  env <- new.env(parent = emptyenv())
  env$label   <- "orders.csv"
  env$columns <- names(starter)
  env$n       <- nrow(starter)
  env$preview <- starter
  env
})
```

`apply_loaded_table()` updates it on every load; each new session calls that
function against it rather than against the starter.

### `querychat` gets the connection, not a data frame

Passing a data frame to `QueryChat$new()` wraps it in a `DataFrameSource` that
freezes rows at construction. Since each session clones its chat client once at
startup, that snapshot would go stale on the next upload. Passing the live
connection builds a `DBISource` whose query tool re-reads the table on every call:

```r
qc <- QueryChat$new(
  dash_con,                    # live connection, not a snapshot
  table_name = "data",
  client = client,
  tools = "query",
  greeting = "greeting.md",
  extra_instructions = build_analyst_prompt(names(starter), nrow(starter),
                                            "orders.csv", df = starter)
)
```

### Swapping providers mid-session

The sidebar switches between the Anthropic and xAI APIs without rebuilding the
querychat module, by replacing the provider inside the live `Chat` object. This
reaches into `ellmer` internals and is version-guarded accordingly:

```r
attach_chat_provider <- function(live_chat, new_chat) {
  if (!ellmer_supported("0.4.2")) {
    warning("ellmer < 0.4.2: in-place provider swap is unsupported.", call. = FALSE)
    return(FALSE)
  }
  tryCatch({
    live_chat$.__enclos_env__$private$provider <- new_chat$get_provider()
    live_chat$set_turns(list())
    TRUE
  }, error = function(e) FALSE)
}
```

If the swap fails the user gets a notification telling them to restart, rather than
a silent fallback to the old model.

| Provider | Models in the picker |
|---|---|
| Anthropic (Claude) | Haiku 4.5 (default), Sonnet 4.6, Sonnet 5 |
| xAI (Grok) | Grok 4.5 (default), Grok 4.6, Grok 4.3 |

Everything shown on this page ran on **Claude Haiku 4.5**. This bills the developer
API, not a Claude.ai or grok.com subscription.

## The code

About 250 KB of R across a dozen files. The chat layer is the thin part.

| File | Lines | Role |
|---|---:|---|
| `R/model-view.R` | 1938 | Method/plot dispatch, spec formatting, metric reporting |
| `R/dashboard-view.R` | 899 | Geom registry, spec resolution, auto-view |
| `R/model-recipe.R` | 782 | Model plot recipes |
| `R/plot-recipe.R` | 766 | One recipe → chart pane and code pane |
| `R/tools.R` | 459 | Typed tool schemas, sticky style merge |
| `R/profile.R` | 451 | Column roles inferred from statistics |
| `R/model-tidymodels.R` | 360 | parsnip / recipes / workflows / yardstick |
| `R/stats-test.R` | 206 | t-test, ANOVA, chi-square, correlation |
| `R/unsupervised.R` | 155 | PCA, k-means, correlation heatmap |
| `R/sql.R` | 156 | DuckDB helpers and the read-only gate |
| `R/providers.R` | 129 | Provider catalog, in-place client swap |
| `app.R` | 1186 | Shiny UI, tool registration, session wiring |

Judgment that would bloat the schema lives in markdown instead:
`extra-instructions.md` covers when to slice versus break down and how to read a
tool return; `ml-instructions.md` covers when to test versus forecast versus lasso.
`greeting.md` and `ml-greeting.md` are static, so startup spends no tokens.

`tests/testthat/` holds 12 files of characterization and agreement tests over the
tool layer — including that the generated code pane agrees with the rendered chart.

## Run it

```r
# 1. Put a key in your user .Renviron (usethis::edit_r_environ()), then restart R:
#      ANTHROPIC_API_KEY=sk-ant-your-key-here
#      XAI_API_KEY=xai-your-key-here

# 2. From the ellmer-practice folder as working directory:
source("setup.R")      # install packages
source("check-key.R")  # optional: prove the key works, for a fraction of a cent
shiny::runApp()
```

$5 of API credit covers a long session. Full setup notes and a troubleshooting
table are in the [repository README](https://github.com/ethandbard/ellmer-practice#readme).

## Limits

- **One browser session.** The DuckDB table is process-wide, so a second tab
  overwrites the first tab's data. This is a workbench, not a multi-user service.
- **Preview fits, not pipelines.** An 80/20 holdout, five-fold CV on the training
  rows, sensible defaults. Enough to see the shape of a problem; not a modeling
  pass you should ship.
- **The gate is a keyword gate.** `is_read_only_sql()` strips literals and comments
  before checking, which handles the obvious evasions, but it is a parser
  approximation rather than a permission boundary. The threat model here is a
  confused model, not a hostile one.
- **The model still chooses.** It writes the SQL and picks the features. Showing
  the query, the spec, and the call is what makes those choices checkable — and as
  the ROC note above shows, the panes catch disagreements that a single number
  would hide.

::: {.page-footer}
Source: [github.com/ethandbard/ellmer-practice](https://github.com/ethandbard/ellmer-practice) ·
Built with R, Shiny, bslib, ellmer, querychat, DuckDB, tidymodels and ggplot2. ·
Screenshots from a live session on `midwest`, August 2026.
:::
